---
title: "A Preliminary Base-case model in SS3.30 for the 2023 North Pacific Swordfish Assessment"
date: "April 2023"
author: "Michelle L Sculley"
format: docx
editor: visual
---

# A preliminary base-case model in SS3.30 for the 2023 North Pacific Swordfish Assessment

Michelle L. Sculley,\* Yi-Jay Chang,^+^ Hirotaka Ijima^\^^

\*NOAA NMFS Pacific Islands Fisheries Science Center, Honolulu, HI

michelle.sculley\@noaa.gov

^+^National Taiwan University, Taipei, Taiwan

^\^^Fisheries Research Agency, Yokohama, Japan

# Abstract

A preliminary base-case model in Stock Synthesis 3.30 for North Pacific (NP) swordfish (*Xiphias gladius*) is described for consideration as the 2023 base-case model. The base-case model covers the Western and Central North Pacific north of the Equator and the Eastern Pacific Ocean north of 10°N from 1975 to 2021. It includes data from three International Scientific Committee for the Conservation of Tuna and Tuna-like Species (ISC) countries and other countries in aggregate from the Western and Central Pacific Fisheries Commission (WCPFC) and the Inter-American Tropical Tuna Commission (IATTC). This paper describes the data available for inclusion and the proposed base-case model. The model converges and appears to fit the data well. Initial diagnostics do not indicate major problems. Preliminary results suggest the North Pacific swordfish stock is being not fished above F~MSY~ and spawning stock biomass is above SSB~MSY~.

# Introduction

The International Scientific Committee for the Conservation of Tuna and Tuna-like Species (ISC) Billfish Working Group (BILLWG) has proposed to run a benchmark assessment on North Pacific swordfish (*Xiphias gladius),* SWO). Data were compiled from the International Scientific Committee for North Pacific Tuna and Tuna-like Species (ISC) member countries and other Western and Central Pacific Fisheries Commission (WCPFC) and Inter-American Tropical Tuna Commission (IATTC) countries. Countries were asked to contribute catch, CPUE, and size-frequency data. It was decided to run the assessment using a sex-specific model in Stock Synthesis version 3.30.20 (Methot and Wetzel, 2013) with fleets-as-areas spearating the Eastern Pacific Ocean (EPO) region from the Western and Central North Pacific (WCNPO) region, as defined in the December 2022 BILLWG meeting. The available data and the preliminary model results and diagnostics for a base-case model will be presented in this document for consideration at the ISC BILLWG SWO stock assessment meeting.

# Methods

## Spatial Temporal Structure

Data were compiled by region assuming a two region model of the North Pacific Ocean with boundaries based upon those detailed in Ichinokawa and Brodziak (2008) with the modification that the Eastern Pacific Ocean (EPO) region ends at 10°N (Figure 1). Countries were asked to contribute catch, CPUE, and length frequency data partitioned by these two regions so that the North Pacific could be modeled with two implicit areas using fleets as areas. The working group agreed to start model in 1975.

## Definition of Fisheries

Data are available for fourteen different fleets in the WCNPO: 14 catch time series, with some fleets split into early and late periods and 8 CPUE indices of which one is a recruitment index. The fleet names and numbers are detailed in Table 1. The data available for each fleet is in Figure 2. The acronyms in the fleet names are defined as follows: WCNPO is Western and Central North Pacific Ocean; EPO is Eastern Pacific Ocean; OSDWLL is offshore distant water longline; OSDWCOLL is offshore distant water and coastal longline; early is the early time period; late is the late time period, Area1 and 2 are the Japanese fishery areas in the WCNPO as defined in Ijima 2022; OSDF is offshore driftnet gear; CODF is coastal driftnet gear, JPN_WCNPO_Other is Japanese small-scale coastal longline vessels which are not under obligation to submit logbook data, bait, and net fishing gear; DWLL is distant water longline gear, TWN_WCNPO_Other is Taiwanese offshore longline, coastal longline, gillnet, harpoon and other gears; LL is longline gear; shallow is the Hawaii shallow-set sector; deep is the Hawaii deep-set sector; GN is gillnet gear; US_WCNPO_Other is harpoon and other gears; WCPFC is other WCPFC and IATTC longline gear in the WCNPO; IATTC is longline gear in the EPO north of 10ºN.

## Catch

Some of the time series of catch were divided into early and late periods to coincide with divisions of the CPUE indices (Table 1, Figure 2). Three ISC countries contributed catch time series: Japan, Taiwan, and the US. In addition, catch from countries reporting to the WCPFC and IATTC were obtained from each RFMO, respectively. The CV for catch was set to 0.05 for all fleets. Catch for fleets with only annual data were divided equally into each quarter.

## Relative Abundance Indices

Each of the eight CPUE time series were assigned to a quarter based upon the recommendations of the country providing the index and are assumed to represent the quarter in which the highest catches take place for each fishery. Japanese longline fleets (S1-4) were all assigned to quarter 1; Taiwanese longline fleet (S5) was assigned to quarter 3; US longline deep-set (S7) was assigned to quarter 2, and US longline shallow-set (S8 and S9) were assigned to quarter 1. US longline deep-set fleet S6 was
included as an index of recruitment because the fishery catches large numbers of young-of-the-year fish (Fleet type 31, Bohaboy and Sculley, 2023). The CPUE indices were assumed to be linearly proportional to biomass where catchability (*q*) was assumed to be constant and occur in the first month of the quarter assigned.

The CVs for each CPUE index were assumed to be equal to their respective calculated SEs on the log scale. The minimum CV was scaled to a minimum of 0.20 or the root-mean-square error (RMSE) (i.e., square root of the residual variance) of what we would expect the assessment model to fit the CPUE index best by adding a constant to each CV value. This was calculated as the square root of the residual variance of a loess smoother fit to each index (Francis 2011, Lee et al., 2014).

\$\$RSME\_{smoother} = \\sqrt{(1/N)\\sum\_{t=1}\^{N}(Y\_{t}-\\hat{Y\_{t}})}\$\$

where *Y*~t~ is the observed CPUE in year *t* on the log scale, $\hat{Y}$ is the predicted CPUE in year *t* from the smoother fit to the data on the log scale, and *N* is the number of CPUE observations. RMSE values for each index are listed in Table 2. If the input SE was greater than these values, it was left unchanged.

## Length Composition

Length composition data were available for six WCNPO fleets and two EPO fleets (Figure 2). Length composition data were available in quarterly time steps. Quarters with fewer than 15 total samples were removed from the time series due to limited sample size, as agreed upon by the modeling sub-group. In addition, the length composition data for F5 were excluded as they were sparse. Data were fit using a multinomial error structure. Length composition data were weighted based upon the Francis (2011) method. Input effective sample size was estimated as the total number of fish measured in each year-quarter divided by 10, with any year-quarter over fifty set to fifty to reduce the influence of large sampling events. Then, weighting was attempted based upon the T.A1.8 equation (Francis 2011) as calculated by the model using r4ss, an R package for plotting SS results (R version 3.4.0, R Core Team, 2017, r4ss version 1.28.0, Taylor et al., 2017). Length composition data were only down-weighted if the model suggested it, otherwise the fleets were not adjusted. The only fleet to be down-weighted using the Francis methods was F9, the US deep-set LL data.

## Initial Base-case Model Description

The assessment was conducted with Stock Synthesis (SS) version 3.30.20.00 released 09/30/2022 using Otter Research ADMB 13.0 (Methot and Wetzel 2013). The model was set up as a single area model with two sexes and four seasons (quarters) with fleets as areas. Spawning was assumed to occur in May (month 5) while recruitment was assumed to occur in July (month 7). Age at recruitment was calculated based upon the model estimated average selectivity at age based upon the quarterly selectivity at length. The maximum age of swordfish was set to 15 years. Sex specific biological parameters were used, with sex- and age-specific natural mortality (Table 3) as agreed upon in the BILLWG data preparatory meeting (ISC Billfish WG 2023). In addition, the CV of the growth curve was set to 0.1 for small males and females and 0.15 for large males and females, and the sex ratio at birth was assumed to be 1:1. The model used a Beverton-Holt spawner-recruit relationship with steepness (h) fixed at 0.9 and sigmaR (σ~r~) initially fixed at 0.6 and rescaled to 0.42 based upon the model suggestion.

Twenty-seven fleets were included in the model: 19 catch fleets and 8 survey fleets. Initial equilibrium catch was estimated for 1974.

Main recruitment deviations were estimated from 1985-2021, as this is the period during which there are size composition data. The recruitment deviations were bias-adjusted based upon the estimates from Methot and Taylor (2011) provided from the model results. No bias adjustment was applied to recruitment deviations from 1965-1967. 1967-1995 was the "ramp-up" period where the bias adjustment of σ~r~ was 0 at the beginning of the period and increased linearly to the maximum bias adjustment 0.94 in 1995. Full bias adjustment was from 1995-2021. The early period of recruitment deviations represents a data-poor period where there is little information to drive recruitment. The main recruitment period represents a data-rich period where there is enough data to drive the bias-adjustment of the recruitments. The ramp up period allows for a gradual ramp up of the bias-adjustment between the data-poor and data-rich periods.

The population model and the fishery length data had 51 five cm length bins from 10-260+ cm. The population had 16 annual ages from age 0 to 15+. There were no age data. Fishery length data were used to estimate selectivity patterns which controlled the size distribution of the fishery removals. All fleets with length data were estimated as six parameter double normal (dome-shaped) selectivity patterns. Survey selectivity patterns mirrored their respective catch fleets (Table 4). Estimating parameters six of the double normal selectivity pattern for F9 resulted in an improved fit to the size composition data for that fleet.

Model estimated time series of total biomass (B in metric tons, mt = 1000 kg), age 1+ total biomass (B~1+~ mt), female spawning biomass (SSB mt) and recruitment (R in 1000s of fish) were tabulated on an annual basis. Annual exploitation rate (F) was calculated as Catch/B~1+~. Stock status indicators were calculated based upon MSY-based reference points as proxies, given that the WCPFC has not set biological or other reference points for swordfish.

## Convergence Criteria and Diagnostics

The model was assumed to have converged if the standard error of the estimated parameters could be derived from the inverse of the negative hessian matrix. Various convergence diagnostics were also evaluated. Excessive CVs (\>50%) on estimated parameters would suggest uncertainty in the parameter estimates or model structure. A gradient of \>0.001 would suggest poorly fit parameter estimates. The correlation matrix was also evaluated to identify highly correlated (\>95%) and non-informative (\<0.01) parameters. Parameter estimates hitting bounds of the prior was also indicative of poor model fit.

Several diagnostics were run to evaluate the fit of the model to the data. An Age-Structure Population Model (APSM) was used to evaluate the influence of the length composition data on the population trends (Carvalho *et al.*, 2017). The ASPM was also used to explore how each CPUE index informed the population trends by running one-off ASPMs for each index. Profiling the likelihood on R~0~, where the R~0~ is fixed at a range ofvalues around the maximum likelihood estimate and then the likelihood is estimated, was used to identify influential data components (Lee *et al.,* 2014). A runs test was used to evaluate randomness in the residuals of the CPUE data (Carvalho *et al.,* 2021). Residual plots and plots of the observed vs expected data were examined to evaluate goodness-of-fit. Finally, a retrospective analysis and hindecast cross-validation were used to evaluate the predictive ability of the model (Carvalho *et al.,* 2021).

# Results

## Model fit

The base-case model ran in about 10 minutes, estimated 89 parameters, and had a total likelihood of 1338.75 The inverse Hessian was positive definite, which allowed for the estimation of parameter standard deviations and suggests that the model converged, and the maximum gradient component was less than 0.001. None of the parameter estimates hit a bound, two selectivity parameters had a correlation above 0.95 and no parameters had correlations below 0.01. All twenty early recruitment deviations (1965-1984) and 30 of the 37 main recruitment deviations had CVs \> 50%. Five of the 34 selectivity parameters had CVs \>50%, most of which were for the estimated width of the peak (i.e. a very small number).

####Fits to the abundance indices were relatively good, with no substantial divergences between the expected and estimated CPUEs (Figures 5 - 10). In addition all indices passed the runs test (Figure 11), which indicates that the residuals are likely random.

Estimated selectivity for each fleet are in Figures 12-15. Fits to the length composition data were also relatively good (Figures 16 - 20), although there are still problems fitting the US longline data (F16).  The fit to the US size data is challenging because it is bimodal, however, current attempts to implement a cubic spline selectivity pattern have not been successful. Furthermore, the estimated mean size of fish caught in the Japanese driftnet fishery is slightly smaller than the observed data (Figure 4). However, all size composition time series passed the runs test (Figure 21).

Model estimates of age 1+ biomass show a slow decrease in biomass from 1975 to 1982, then biomass varied around MSY, declined to its lowest level in 1998, and has relatively stable since (Figure 22). Initial spawning stock biomass was estimated to be approximately 4,200 mt and virgin SSB was around 25,000 mt (Figure 23). Annual fishing mortality is reported as the average for fish ages 3-12 (Figure 24). Fishing mortality was above MSY for all except 6 years and excepting 2015, has been below F~MSY~ since 2014. Recruitment deviations suggested three periods of recruitment: high recruitment from 1975 to 1993, a period with little data and large variability around the estiamtes, average recruitment from 1994 to 2003 where recruitment varied around equilibrium recreuitment, and low recruitment from 2004 to 2020. The log of the deviations were generally between 0.6 and -0.6 (Figure 25). Current depletion, as estimated as the age 1+ biomass in 2020 compared to the virgin age 1+ biomass was estimated to be 0.09.

## Diagnostics

Profiling on R~0~ showed that the recruitment estimates were highly influential in the model results, and there was substantial conflict between the CPUE indices and the length composition data (Figures 26-28). The US data (CPUE and length comp) drive the model dynamics suggesting an ln(R~0~) below 6.0, and Chinese Taipei data and Japanese size composition data suggesting an ln(R~0~) around 7. Japanese CPUE data suggest an ln(R~0~) around 6.1 (Tables 4-5).

Results from the ASPM model showed the same population trend as the full model during 1975-1993, which is the time period without any size composition data. After 1994, the ASPM biomass increases drastically and deviates completely from the base-case model (Figures 29 and 30). Further investigation running an ASPM with a single early and late index at a time indicates a similar pattern for all CPUE indices (Figure 31). The most likely explanation for this is that the population dynamics prior to 1994 are being driven by the two CPUE indices available (Figure 32). After 1994 catch decreases and CPUE for all fleets flattens out which would indicate that a stock is recovering. However, the size composition data indicate that the majority of the catch is juvenile fish. Continued removals of individuals before they have a chance to reproduce would continue to cause the stock to decline below MSY levels. This indicates that the size composition data are an integral component of the model, without which we would not have a full picture of the fishing effect on the stock.

The retrospective analysis indicates that a significant retrospective pattern exists for both biomass and fishing mortality (Mohn's rho = 0.2 and -0.14, respectively, Figure 33). Generally, biomass is overestimated and fishing mortality is underestimated. Results of the hindcast with cross-validation indicate that of the four CPUE indices at the end of the assessment horizon, only Chinese Taipei had reasonable predictive ability (MASE = 0.9), with all other fleets MASE \> 1 (Figure 34). Comparing the predictive ability of the size composition data, two fleets had very good predictive ability (MASE \<0.5, F2 and F18), five had good predictive ability (MASE \<1 and \>0.5, F1, F4, F5, F6, F16) and one had poor predictive ability (MASE \<1, F14, Figure 35). F14 only had one datapoint in the analysis which likely explains why it had poor predictive ability.

The likelihood profile indicated that the US data componenets (CPUE and size composition data) are majority contributors to the likelihood. Attempts to downweight the US LL size composition data results in a model that fails to converge (due to a Hessian that is not positive definite). However, the ASPM models indicate that the US CPUE data suggest a similar trend as the other CPUE indices. To further explore the effect of the US data, models were run removing all US data except catch and fitting the parameters. While the models including the Japanese size data also failed to converge, the results for all the models indicated an estimated ln(Ro) between 6.2 and 6.4, and SSB~zero~ between 12500 and 15500 metric tons (Table 6). This suggests that even though the US data is a significant contributor to the likelihood, the estimated population size would be similar without the data.

Compared to 2019 base-case model, the 2022 base-case model spawning stock biomass estimates are larger and decline at the start of the assessment (Figure 37). However, the trend and SSB/SSBMSY after 1993 are fairly similar. The main driver of the difference at the beginning of the assessment is that the initial fishing mortality is estimated in 2022 but fixed in 2019, and Japanese driftnet catches have been corrected to lower values in 2022. Fishing mortality is less variable in the 2022 model compared to the 2019 model and is generally lower in 1993 to 2020, with the last five years of the 2022 model below F~MSY~ (Figure 38). F~MSY~ was estimated to be approximately equal for both models (0.63 for the 2022 model and 0.61 for the 2019 model. Recruitment deviations are very similar between the two models except in the very beginning of the model, but this is a period with very high variability (Figure 39).

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
